Gravitational-wave sensitivity curves

### 2014-08-06 Detectors.tex
Post-arXiv 1

In this section we introduce the detector noise curves used in \ref{app:a}. We begin with a description of the basic operation of detectors. We then discuss ground-based detectors, space-based detectors and PTAs in turn. The latter function somewhat differently than conventional interferometers, so we include a brief introduction to PTA analysis. References for the noise curves used for individual detectors can be found in the relevant subsections, further information about the detectors can be in discovered these. Detectors are frequently upgraded and redesigned, hence, while these curves are believed to be correct at the time of writing, it is best to check for updates from the appropriate science teams before relying on the details given here, although we hope that they shall remain accurate enough for illustrative purposes.

@@

The response function of a two-arm interferometric detector is quadrupolar, an example is plotted in figure \ref{fig:LIGO}.
\begin{figure}
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.99\textwidth]{LIGO_detectorframe.pdf}
 \caption{The angular response function of an interferometric detector shown both as a surface plot and in an Aitoff--Hammer projection. The quentity that is plotted is is the polarisation average, $\int\textrm{d}\psi \;((F^{+})^{2}+(F^{\times})^{2})/2$. The response is a function of two sky angles, $\theta$ and $\phi$, and varies between $0$ and $1$. The two detector arms lie in the $x$--$y$ plane either side of one of the zeros in the response.}
 \label{fig:LIGO}
\end{figure}
Throughout this paper detector sensitivity refers to the polarisation and sky averaged sensitivity $F$, where
\begin{equation}\label{eq:skyav}
F^{2}=\int_{0}^{2\pi}\frac{\mathrm{d}\psi}{2\pi}\; \int_{0}^{2\pi} \frac{\mathrm{d}\phi}{2\pi}\; \int_{0}^{\pi}\frac{\sin\theta\,\mathrm{d}\theta}{2}\;\left[\frac{F^{+}\left(\theta,\phi,\psi\right)^{2}+F^{\times}\left(\theta,\phi,\psi\right)^{2}}{2}\right]\; .
\end{equation}
For a single $90\degree$-interferometer, such as LIGO, the sky and polarisation averaged response is $F=\sqrt{1/5}\approx 0.447$.

@@

The most notable ground-based detectors are LIGO and Virgo, which work in collaboration, supported by GEO600. LIGO, Virgo and GEO600 have completed science runs as first-generation detectors. As LIGO and Virgo are currently being upgraded their initial configurations are now referred to as Initial LIGO (iLIGO) and Initial Virgo (iVirgo) respectively. The upgraded, second-generation versions are referred to as advanced LIGO (aLIGO) and advanced Virgo (AdV) respectively. LIGO has two observatories: one at Hanford, Washington, which has two detectors; and another at Livingston, Louisiana. There is an agreement to move one of the upgraded Hanford detector systems to a location in India \citep{LIGO-India,Unnikrishnan2013}. The GEO600 detector can also operate in a high frequency mode \citep[GEO-HF]{2006CQGra..23S.207W}. The advanced detectors should start operation in the next couple of years, with LIGO-India following further in the future.

@@

Space-based detectors work on similar principles to ground based detectors, but with the test masses residing inside of independent, widely separated satellites. Space-based detectors are sensitive to lower frequency GWs than their ground-based counterparts; this is partly because space based detectors can have much longer arms, and partly because they are unaffected by seismic noise which limits the low frequency performance of ground-based detectors.

@@

\Gamma_{0} = \frac{1}{2}+\frac{3x}{2}\left(\ln x -\frac{1}{6}\right) \; ,

@@

where $x = (1-\cos\theta)/{2}.$

@@

 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.6\textwidth]{HnDcurve.pdf}

@@

There is a discrepancy between the treatment of PTA sensitivity curves here and the higher frequency detectors discussed in sections \ref{sec:ground} and \ref{sec:space}. When observing a long-lived source, such as an inspiral, with a high frequency detector the convention was to define a \emph{characteristic} strain to satisfy (\ref{eq:hc}). Here, the convention is to leave the strain untouched and instead adjust the PTA sensitivity curve with observation time, again to satisfy (\ref{eq:hc}). This discrepancy is an unfortunate result of the conventions in use by the different GW communities; however, it is also natural given the sources under observation. When observing a transient source, such as a burst or inspiral, which changes within the lifetime of the detector, it is natural to consider the detector as performing constantly while the signal changes. However, when observing a monochromatic source or a stochastic background, which is unchanging over the detector lifetime, it is more natural to consider the source as being fixed and the sensitivity of the detector gradually improving. All that is required by the definition in (\ref{eq:hc}) is that the ratio $h_\mathrm{c}(f)/h_{n}(f)$ is constant.

@@

The PTAs currently in operation are the European Pulsar Timing Array \citep[EPTA\footnote{\url{http://www.epta.eu.org/}};][]{eptareview2013}, the Parkes Pulsar Timing Array \citep[PPTA\footnote{\url{http://www.atnf.csiro.au/research/pulsar/ppta/}};][]{parkesreview2013} based in Australia, and the North American Nanohertz Observatory for Gravitational waves \citep[NANOGrav\footnote{\url{http://nanograv.org/}};][]{nanogravreview2013}). There are published limits on the amplitude of the stochastic background from all three detectors: the most recent from EPTA is \citet{Haasteren}, from PPTA is \citet{Shannon2013} and from NANOGrav is \citet{2013ApJ...762...94D}. These are all comparable and we use the EPTA limits in \ref{app:a}, the figures show the published limit on $A$ from equation \ref{eq:power} at a frequency of $f_{0}=1\,\textrm{year}^{-1}$, based on an analysis of five pulsars over approximately ten years.

@@

Combining the existing arrays would yield a single PTA using approximately 4 times as many pulsars; this consortium of consortia is known as the International Pulsar Timing Array \citep[IPTA\footnote{\url{http://www.ipta4gw.org/}};][]{iptareview2013}. The IPTA curves plotted in the figures are based on 20 pulsars timed for twice as long as the current EPTA.

@@

The next great advancement in radio astronomy shall come with the completion of the Square Kilometre Array \citep[SKA;][]{Dewdney2009}. This shall greatly increase the sensitivity of pulsar timing \citep{Kramer2004}. The sensitivity curves plotted in \ref{app:a} for the SKA assume a root-mean-square error on the timing residuals a factor of 10 better than current PTAs, and include 20 pulsars timed for twice as long as the current EPTA.

>>>>

In this section we introduce the detector noise curves used in \ref{app:a}. We begin with a description of the basic operation of detectors. We then discuss ground-based detectors, space-based detectors and PTAs in turn. The latter function somewhat differently than conventional interferometers, so we include a brief introduction to PTA analysis. References for the noise curves used for individual detectors can be found in the relevant subsections and further information about the detectors can be discovered in these. Detectors are frequently upgraded and redesigned, hence, while these curves are believed to be correct at the time of writing, it is best to check for updates from the appropriate science teams before relying on the details given here, although we hope that they shall remain accurate enough for illustrative purposes.

@@

The response function of a two-arm interferometric detector is quadrupolar, an example is plotted in figure \ref{fig:LIGO}.
\begin{figure}
 \centering
 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.99\textwidth]{LIGO_detectorframe.pdf}
 \caption{The angular response function of an interferometric detector shown both as a surface plot and in an Aitoff--Hammer projection. The quentity that is plotted is is the polarisation average, $\int\textrm{d}\psi \;((F^{+})^{2}+(F^{\times})^{2})/2$ \cplb{Should this be divided by $2\pi$ too?}. The response is a function of two sky angles, $\theta$ and $\phi$, and varies between $0$ and $1$. The two detector arms lie in the $x$--$y$ plane either side of one of the zeros in the response.}
 \label{fig:LIGO}
\end{figure}
Throughout this paper detector sensitivity refers to the polarisation and sky averaged sensitivity $F$, where
\begin{equation}\label{eq:skyav}
F^{2}=\int_{0}^{2\pi}\frac{\mathrm{d}\psi}{2\pi}\; \int_{0}^{2\pi} \frac{\mathrm{d}\phi}{2\pi}\; \int_{0}^{\pi}\frac{\sin\theta\,\mathrm{d}\theta}{2}\;\left[\frac{F^{+}\left(\theta,\phi,\psi\right)^{2}+F^{\times}\left(\theta,\phi,\psi\right)^{2}}{2}\right]\; .
\end{equation}
For a single $90\degree$-interferometer, such as LIGO, the sky and polarisation averaged response is $F=\sqrt{1/5}\approx 0.447$.

@@

The most notable ground-based detectors are LIGO and Virgo, which work in collaboration, supported by GEO600. LIGO, Virgo and GEO600 have completed science runs as first-generation detectors \cplb{[citation needed]}. As LIGO and Virgo are currently being upgraded their initial configurations are now referred to as Initial LIGO (iLIGO) and Initial Virgo (iVirgo) respectively. The upgraded, second-generation versions are referred to as Advanced LIGO (aLIGO) and Advanced Virgo (AdV) respectively. LIGO has two observatories: one at Hanford, Washington, which has two detectors; and another at Livingston, Louisiana. There is an agreement to move one of the upgraded Hanford detector systems to a location in India \citep{LIGO-India,Unnikrishnan2013}. The GEO600 detector can also operate in a high frequency mode \citep[GEO-HF]{2006CQGra..23S.207W} \cplb{Add to table}. The advanced detectors should start operation in the next couple of years, with LIGO-India following further in the future.

@@

Space-based detectors work on similar principles to ground based detectors, but with the test masses residing inside of independent, widely separated satellites. Space-based detectors are sensitive to lower frequency GWs than their ground-based counterparts; this is partly because space-based detectors can have much longer arms, and partly because they are unaffected by seismic noise which limits the low frequency performance of ground-based detectors.

@@

\Gamma_{0} = \frac{1}{2}+\frac{3\xi}{2}\left(\ln \xi -\frac{1}{6}\right) \; ,

@@

where $\xi = (1-\cos\theta)/{2}.$

@@

 \includegraphics[trim=0cm 0cm 0cm 0cm, width=0.6\textwidth]{Fig_Hellings_Downs}

@@

There is a discrepancy between the treatment of PTA sensitivity curves here and the higher frequency detectors discussed in sections \ref{sec:ground} and \ref{sec:space}. When observing a long-lived source, such as an inspiral, with a high frequency detector the convention was to define a \emph{characteristic} strain to satisfy (\ref{eq:hc}). Here, the convention is to leave the strain untouched and instead adjust the PTA sensitivity curve with observation time, again to satisfy (\ref{eq:hc}). This discrepancy is an unfortunate result of the conventions in use by the different GW communities; however, it is natural given the sources under observation. When observing a transient source, such as a burst or inspiral, which changes within the lifetime of the detector, it is natural to consider the detector as performing constantly while the signal changes. However, when observing a monochromatic source or a stochastic background, which is unchanging over the detector lifetime, it is more natural to consider the source as being fixed and the sensitivity of the detector gradually improving. All that is required by the definition in (\ref{eq:hc}) is that the ratio $h_\mathrm{c}(f)/h_{n}(f)$ is constant.

@@

The PTAs currently in operation are the European Pulsar Timing Array \citep[EPTA\footnote{\url{http://www.epta.eu.org/}};][]{eptareview2013}, the Parkes Pulsar Timing Array \citep[PPTA\footnote{\url{http://www.atnf.csiro.au/research/pulsar/ppta/}};][]{parkesreview2013} based in Australia, and the North American Nanohertz Observatory for Gravitational waves \citep[NANOGrav\footnote{\url{http://nanograv.org/}};][]{nanogravreview2013}. There are published limits on the amplitude of the stochastic background from all three detectors: the most recent from EPTA is \citet{Haasteren}, from PPTA is \citet{Shannon2013} and from NANOGrav is \citet{2013ApJ...762...94D}. These are all comparable and we use the EPTA limits in \ref{app:a}, the figures show the published limit on $A$ from equation \ref{eq:power} at a frequency of $f_{0}=1~\mathrm{yr^{-1}}$, based on an analysis of $5$ pulsars over approximately $10~\mathrm{yr}$.

@@

Combining the existing arrays would yield a single PTA using approximately $4$ times as many pulsars; this consortium of consortia is known as the International Pulsar Timing Array \citep[IPTA\footnote{\url{http://www.ipta4gw.org/}};][]{iptareview2013}. The IPTA curves plotted in the figures are based on $20$ pulsars timed for twice as long as the current EPTA.

@@

The next great advancement in radio astronomy shall come with the completion of the Square Kilometre Array \citep[SKA;][]{Dewdney2009}. This shall greatly increase the sensitivity of pulsar timing \citep{Kramer2004}. The sensitivity curves plotted in \ref{app:a} for the SKA assume a root-mean-square error on the timing residuals a factor of $10$ better than current PTAs, and include $20$ pulsars timed for twice as long as the current EPTA.

### 2014-08-06 Introduction.tex
Post-arXiv 1

The next few years  promise to deliver the first direct detection of gravitational waves (GWs). This will most likely be achieved by the advanced Laser Interferometer Gravitational-wave Observatory \citep[LIGO;][]{2010CQGra..27h4006H} and advanced Virgo \citep{Acernese2009} detectors operating in the frequency range $(10$--$10^{3})~\mathrm{Hz}$. By the end of the decade, it is expected that pulsar timing arrays \citep[PTAs;][]{1990ApJ361300F} will also detect very low frequency GWs around $10^{-8}~\mathrm{Hz}$. Further into the future, space-based detectors, such as the evolved Laser Interferometer Space Antenna \citep[eLISA;][]{TheGravitationalUniverse}, will probe GWs in the millihertz regime. These advances shall herald the beginning of multi-wavelength GW astronomy as a means of observing the Universe.

@@

We begin by expounding the various conventions and the relationships between the conventions in sections \ref{sec:conventions} and \ref{sec:voc}. A review of GW detectors (both current and proposed) is given in section \ref{sec:detectors} and a review of GW sources is given in section \ref{sec:sources}. In \ref{app:a} several example sensitivity curves are presented. A website where similar figures can be generated is available at \url{www.ast.cam.ac.uk/~rhc26/sources/}. Here, the user may select which sources and detectors to include to tailor the figure to their specific requirements.

@@

Since the GW signal and detector output are both real it follows that $\tilde{h}(-f)=\tilde{h}^{*}(f)$ and $\tilde{n}(-f)=\tilde{n}^{*}(f)$; therefore, $S_{n}(f)=S_{n}(-f)$. The fact that $S_{n}(f)$ is an even function means that Fourier integrals over all frequencies can instead be written as integrals over positive frequencies only, e.g., (\ref{eq:meansquare1}) and (\ref{eq:snrinnerprod}); it is for this reason that $S_{n}(f)$ is called the \emph{one-sided} PSD.\footnote{An alternative convention is to use the two-sided PSD $S^{(2)}_{n}(f) = S_{n}(f)/2$.}

When integrated over all positive frequencies, the PSD gives the mean square noise amplitude. Starting by taking the time average of the square of the detector noise:
\begin{eqnarray}
\overline{\left|n(t)\right|^{2}} &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; n(t)n^{*}(t) \\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\tilde{n}(f)\tilde{n}^{*}(f')\exp\left(2\pi \rmi ft\right)\exp\left(-2\pi \rmi f't\right) \nonumber\\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\mathcal{F}}\left\{ n(\tau) \right\}(f)\left[{\mathcal{F}}\left\{ n(\tau) \right\}(f')\right]^{*} \nonumber \\*
\label{eq:meansquare} & & \times \hphantom{\int_{-T}^{T}} \exp\left(2\pi \rmi ft\right)\exp\left(-2\pi \rmi f't\right)\;,
\end{eqnarray}
where we have substituted in using the definitions of the Fourier transform and its inverse. A property of Fourier transforms is that a time-domain translation by amount $t$ is equivalent to a frequency-domain phase change $2\pi ft$; if ${\mathcal{F}}\left\{ n(\tau) \right\}(f) = \tilde{n}(f)$, then ${\mathcal{F}}\left\{ n(\tau-t) \right\}(f) = \tilde{n}(f)\exp (2\pi \rmi ft )$. Therefore, the exponential factors in (\ref{eq:meansquare}) may be absorbed as
\begin{eqnarray}
\overline{\left|n(t)\right|^{2}} &=& \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T} \mathrm{d}t\:\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\mathcal{F}}\left\{ n(\tau-t) \right\}(f)\left[{\mathcal{F}}\left\{ n(\tau-t) \right\}(f')\right]^{*} \, .
\end{eqnarray}
Since the noise is a randomly varying signal, we can use the ergodic principle to equate a time average, denoted by $\overline{\left(\ldots\right)}$, with an ensemble average, denoted by $\left<\ldots\right>$. The noise is stationary, consequently, its expectation value is unchanged by the time-translation performed above. Therefore, using (\ref{eq:psd}), the mean square noise amplitude is given by
\begin{eqnarray}\label{eq:meansquare1}
\overline{\left|n(t)\right|^{2}} &=& \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right> \\
&=& \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\frac{1}{2}S_{n}(f)\delta(f-f') \nonumber \\
&=& \int_{0}^{\infty}\mathrm{d}f\; S_{n}(f) \;.
\end{eqnarray}

@@

Whilst the magnitude of the Fourier transform of the signal $|\tilde{h}(f)|$ provides a simple quantification of the GW amplitude as a function of frequency, it has one main deficiency. For an inspiralling source ,the instantaneous amplitude can be orders of magnitude below the noise level in a detector; however, as the signal continues over many orbits, the SNR can be integrated up to a detectable level. It is useful to have a quantification of the GW amplitude that accounts for this effect; three such methods are described in the following subsections.

@@

The PSD, as defined by (\ref{eq:psd}), has the nice property, demonstrated in (\ref{eq:meansquare1}), that integrated over all positive frequencies it gives the mean square amplitude of the signal in the detector. However, in one important regard it is less appealing than characteristic strain: the height of the source above the detector curve is no longer directly related to the SNR.

@@

This quantity has one aesthetic advantage over the others: it automatically accounts for the fact that there is less energy in low frequency waves of the same amplitude. However, unlike characteristic strain, the area between the source and detector curves is no longer simply related to the SNR.

\subsection{Relating the different descriptions}
The dimensionless energy density in GWs $\Omega_{\mathrm{GW}}$, spectral energy density $S_{\mathrm{E}}$, one-sided PSD, $S_{h}$, characteristic strain $h_\mathrm{c}$ and frequency-domain strain $\tilde{h}(f)$ are related via
\begin{equation}\label{eq:differentdescriptions}
H_0^2\Omega_\mathrm{GW}(f)= \frac{8 \pi G}{3 c^{2}} fS_{\mathrm{E}}(f) = \frac{2\pi^2}{3} f^3 S_h(f) = \frac{2\pi^2}{3} f^2 h_\mathrm{c}^2(f) = \frac{8\pi^2}{3} f^4 \left|\tilde{h}(f)\right|^2\; ,
\end{equation}
using (\ref{eq:strain-hc}), (\ref{eq:ShforSources}), (\ref{eq:spectralenergydensity}), (\ref{eq:omega}) and (\ref{eq:crit-density}).
Corresponding expressions for the noise are obtained by substituting $S_{n}(f)$ for $S_h(f)$, $h_{n}(f)$ for $h_\mathrm{c}(f)$ and $\tilde{n}(f)$ for $\tilde{h}(f)$.

>>>>

The next few years  promise to deliver the first direct detection of gravitational waves (GWs). This will most likely be achieved by the advanced versions of the Laser Interferometer Gravitational-wave Observatory \citep[LIGO;][]{2010CQGra..27h4006H} and Virgo \citep{Acernese2009} detectors operating in the frequency range $(10$--$10^{3})~\mathrm{Hz}$. By the end of the decade, it is expected that pulsar timing arrays \citep[PTAs;][]{1990ApJ361300F} will also detect very low frequency GWs around $10^{-8}~\mathrm{Hz}$. Further into the future, space-based detectors, such as the evolved Laser Interferometer Space Antenna \citep[eLISA;][]{TheGravitationalUniverse}, will probe GWs in the millihertz regime. These advances shall herald the beginning of multi-wavelength GW astronomy as a means of observing the Universe.

@@

We begin by expounding the various conventions and the relationships between the conventions in sections \ref{sec:conventions} and \ref{sec:voc}. A review of GW detectors (both current and proposed) is given in section \ref{sec:detectors} and a review of GW sources is given in section \ref{sec:sources}. In \ref{app:a}, several example sensitivity curves are presented. A website where similar figures can be generated is available at \url{www.ast.cam.ac.uk/~rhc26/sources/}. Here, the user may select which sources and detectors to include to tailor the figure to their specific requirements.

@@

Since the GW signal and detector output are both real it follows that $\tilde{h}(-f)=\tilde{h}^{*}(f)$ and $\tilde{n}(-f)=\tilde{n}^{*}(f)$; therefore, $S_{n}(f)=S_{n}(-f)$. The fact that $S_{n}(f)$ is an even function means that Fourier integrals over all frequencies can instead be written as integrals over positive frequencies only, e.g., (\ref{eq:meansquare2}) and (\ref{eq:snrinnerprod}); it is for this reason that $S_{n}(f)$ is called the \emph{one-sided} PSD.\footnote{An alternative convention is to use the two-sided PSD $S^{(2)}_{n}(f) = S_{n}(f)/2$.}

When integrated over all positive frequencies, the PSD gives the mean square noise amplitude. Starting by taking the time average of the square of the detector noise:
\begin{eqnarray}
\overline{\left|n(t)\right|^{2}} &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; n(t)n^{*}(t) \\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\tilde{n}(f)\tilde{n}^{*}(f')\exp\left(2\pi \rmi ft\right)\exp\left(-2\pi \rmi f't\right) \nonumber\\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\mathcal{F}}\left\{ n(\tau) \right\}(f)\left[{\mathcal{F}}\left\{ n(\tau) \right\}(f')\right]^{*} \nonumber \\*
\label{eq:meansquare} & & \times \vphantom{\int_{-T}^{T}} \exp\left(2\pi \rmi ft\right)\exp\left(-2\pi \rmi f't\right)\;,
\end{eqnarray}
where we have substituted in using the definitions of the Fourier transform and its inverse. A property of Fourier transforms is that a time-domain translation by amount $t$ is equivalent to a frequency-domain phase change $2\pi ft$; if ${\mathcal{F}}\left\{ n(\tau) \right\}(f) = \tilde{n}(f)$, then ${\mathcal{F}}\left\{ n(\tau-t) \right\}(f) = \tilde{n}(f)\exp (2\pi \rmi ft )$. Therefore, the exponential factors in (\ref{eq:meansquare}) may be absorbed as
\begin{eqnarray}
\overline{\left|n(t)\right|^{2}} &=& \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T} \mathrm{d}t\:\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\mathcal{F}}\left\{ n(\tau-t) \right\}(f)\left[{\mathcal{F}}\left\{ n(\tau-t) \right\}(f')\right]^{*} \, .
\end{eqnarray}
Since the noise is a randomly varying signal, we can use the ergodic principle to equate a time average, denoted by $\overline{\left(\ldots\right)}$, with an ensemble average, denoted by $\left<\ldots\right>$. The noise is stationary, consequently, its expectation value is unchanged by the time-translation performed above. Therefore, using (\ref{eq:psd}), the mean square noise amplitude is given by
\begin{eqnarray}\label{eq:meansquare1}
\overline{\left|n(t)\right|^{2}} &=& \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right> \\
&=& \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\frac{1}{2}S_{n}(f)\delta(f-f') \nonumber \\
&=& \int_{0}^{\infty}\mathrm{d}f\; S_{n}(f) \;.
\label{eq:meansquare2}
\end{eqnarray}

@@

Whilst the magnitude of the Fourier transform of the signal $|\tilde{h}(f)|$ provides a simple quantification of the GW amplitude as a function of frequency, it has one main deficiency. For an inspiralling source, the instantaneous amplitude can be orders of magnitude below the noise level in a detector; however, as the signal continues over many orbits, the SNR can be integrated up to a detectable level. It is useful to have a quantification of the GW amplitude that accounts for this effect; we shall now describe how this can be achieved.

@@

The PSD, as defined by (\ref{eq:psd}), has the nice property, demonstrated in (\ref{eq:meansquare2}), that integrated over all positive frequencies it gives the mean square amplitude of the signal in the detector. However, in one important regard it is less appealing than characteristic strain: the height of the source above the detector curve is no longer directly related to the SNR.

@@

This quantity has one aesthetic advantage over the others: it automatically accounts for there being less energy in low frequency waves of the same amplitude. However, unlike characteristic strain, the area between the source and detector curves is no longer simply related to the SNR.

\subsection{Relating the different descriptions}
The dimensionless energy density in GWs $\Omega_{\mathrm{GW}}$, spectral energy density $S_{\mathrm{E}}$, one-sided PSD, $S_{h}$, characteristic strain $h_\mathrm{c}$ and frequency-domain strain $\tilde{h}(f)$ are related via
\begin{equation}\label{eq:differentdescriptions}
H_0^2\Omega_\mathrm{GW}(f)= \frac{8 \pi G}{3 c^{2}} fS_{\mathrm{E}}(f) = \frac{2\pi^2}{3} f^3 S_h(f) = \frac{2\pi^2}{3} f^2 \left[h_\mathrm{c}(f)\right]^2 = \frac{8\pi^2}{3} f^4 \left|\tilde{h}(f)\right|^2\; ,
\end{equation}
using (\ref{eq:strain-hc}), (\ref{eq:ShforSources}), (\ref{eq:spectralenergydensity}), (\ref{eq:omega}) and (\ref{eq:crit-density}).
Corresponding expressions for the noise are obtained by substituting $S_{n}(f)$ for $S_h(f)$, $h_{n}(f)$ for $h_\mathrm{c}(f)$ and $\tilde{n}(f)$ for $\tilde{h}(f)$.
